---
title: "A5_phylogeny_report"
author: "J Moggridge"
date: "10/12/2020"
output: pdf_document
highlight: kate
urlcolor: blue
---

This work is available on [github](https://github.com/jmoggridge/6210_project5_phylogeny)

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
options(tinytex.verbose = TRUE)
```


```{r}
# Libraries
library(tidyverse)
library(bold)
library(Biostrings)
library(DECIPHER)
library(beepr)
library(rcartocolor)
library(ggthemes)
library(patchwork)
```


### 1. Introduction

<!-- 
Paragraph 1: What is the overarching topic you are interested in? Why is this scientifically important and interesting and/or societally relevant?o
 -->
Polychaeta are diverse group... many cryptic species..

Metabarcoding is commonly used as a means for biodiversity analysis; these are used for ecological assessments and environmental monitoring.

Climate change is a major threat to global biodiversity, however the effects upon the arctic region is been unparalleled.

Phylogenetic analysis are important for understanding the diversity of species in a community.


<!-- dkd
Paragraph 2: You can outline what is an important sub-area of research or a gap in knowledge. (As an example, perhaps your broader theme is gene expression analysis, and in paragraph 2 you could narrow in to introduce the importance of statistical choices.)
-->
Model choice remains a topic of contention (Abad 2019i; Crandall; Posata and Crandall).

As there is no way to ascertain the actual course of substitutions, we must apply biases (eg. parsimony, likelihood, information criteria) to even have a grounds upon which to evaluate different models...

Perhaps a holistic view considers various models as an ensemble.

Polynoidae were selected as a convienent study subject having ample data on BOLD, with a non-trivial number of genera and species.
<!--
Paragraph 3: What is the specific objective of your study? What kind of study are you performing? What will you d 
-->


### 2. Data

barcode sequences were downloaded from BOLD on 2020-12-01.


```{r eval=FALSE}
 # BOLD downlaod for taxon 'Polychaeta'
polynoid.bold <- bold::bold_seqspec(taxon = 'Polynoidae')
write_rds(polynoid.bold, 'polynoidea.bold.rds', compress = 'gz')
```



